\textit{Supporting Information for}: Geometric models reveal behavioral and neural signatures of how naturalistic experiences are transformed into episodic memories

### 2020-08-20 paper/supporting_information.tex
updated video -> episode in supplement

\begin{figure}[]
\centering
\includegraphics[width=1\textwidth]{figs/feature_value}
\caption{\small \textbf{Feature importance analysis.} \textbf{A.} Contributions of each feature type to the structure of the video trajectory. The bar heights reflect the correlation between the video trajectory computed using all features with a video trajectory computed using all features except the indicated feature.  (Lower bars reflect features that contribute more substantially to the video trajectory's shape.) \textbf{B.} Which features are preserved during recall?  The bar heights reflect the (average) across-participant correlations between the video and recall trajectories.  Error bars denote bootstrap-estimated standard error of the mean.  \textbf{C.} Feature correlation matrix.  Each entry displays the correlation between video topic trajectories created using only the indicated (row/column) features.}
\label{fig:feature-importance}
\end{figure}

We also carried out an analysis of which annotated features tended to shape aspects of the video topic trajectory's structure that were preserved in participants' recalls.  Specifically, we computed the timepoint-by-timepoint correlation matrix of the video topic trajectory, and correlated the proximal diagonals from its upper triangle with those of the timepoint-by-timepoint correlation matrices for each participant's recall topic trajectory (resampled using linear interpolation to have the same number of timepoints as the video trajectory).  This yielded a single correlation coefficient for each participant.  We then repeated this analysis with each annotated feature held out in turn.  Observing a lower correlation between the video and recall trajectories (when a given feature was held out) would indicate that participants utilize changes in that feature's content to discriminate between sections of the video when organizing their recalls.  We found that hand-annotated narrative details were the most heavily utilized type of feature, whereas changes in the text present on-screen, the indoor/outdoor distinction between scenes, the camera angle, the names of the various characters on screen, and which character was speaking tended not to impact participants' recall structures (i.e., removing those features resulted in a \textit{greater} video-recall topic vector correlation than including them; Fig.~\ref{fig:feature-importance}B).

Next, we wondered how the different types of features might relate.  For example, knowing which character is in focus during a given scene may also provide information about which character is speaking.  We computed video topic trajectories from annotations of each individual feature, in turn, and then compared the proximal diagonals from the upper triangles of the temporal correlation matrices of all pairs of features.  This provided additional confirmation that the shape of the full trajectory (including all types of features) was largely driven by narrative details.  We also found that character-driven features (characters on screen, characters speaking, and characters in focus) were strongly correlated.  Other details, such as the presence or absence of music, led to very different topic trajectories (Fig.~\ref{fig:feature-importance}C).
%\FloatBarrier

\subsection*{Creating a low-dimensional embedding space}
Figures~7 and 8C in the main text display two-dimensional projections of the 100-dimensional topic trajectories for the video (7A, 8C), average recall (7B), and each individual's recall (7C, 8C).  We created these embeddings using the Uniform Manifold Approximation and Projection algorithm \citep[UMAP;][]{McInEtal18} called from our high-dimensional visualization and text analysis software, \texttt{HyperTools}~\citep{HeusEtal18a}.  An advantage of the UMAP algorithm over comparable manifold learning techniques (such as $t$-SNE) is that UMAP explicitly attempts to preserve the global structure of the data \citep{McInEtal18,BechEtal19} by constructing a space where distance on the manifold is standard Euclidean distance, with respect to the global coordinate system.  This was important in our use case, as we wanted to visualize both the evolving structure of the video and the spatial relationships between presented and recalled content.

@@

Two other parameters ultimately affect this balance between preserving local and global structure: the seed ($\tau$) for the (pseudo-)random number generator, and the order of the observations (i.e., trajectories) to be embedded ($\varphi$).  As described in \textit{Methods}, we ensured the video and recall events were projected onto the \textit{same} low-dimensional manifold by fitting the embedding model to a stacked matrix of all video, average recall, and individuals' recall events.  After initializing the low-dimensional simplicial set (by default, using a spectral embedding of the high-dimensional simplicial set's fuzzy 1-skeleton), UMAP optimizes the embedding using stochastic gradient descent with cross-entropy as a cost function.  During optimization, indices of the data are sampled at random, and thus the local minimum achieved by the optimization is dependent on both the state of the RNG and the sequence in which observations are concatenated.

We performed a grid search over pre-specified values of these hyper-parameters, and optimized the manifold space for visualization based on two criteria: First, that the 2D embedding of the video trajectory should reflect its original 100-dimensional structure as faithfully as possible. Second, that the path traversed by the embedded video trajectory should intersect itself a minimal number of times.  The first criteria helps bolster the validity of visual intuitions about relationships between sections of video content, based on their locations in the embedding space.  The second criteria was motivated by the observed low off-diagonal values in the video trajectory's temporal correlation matrix (suggesting that the same topic-space coordinates should not be revisited; see Figure~2A in the main text) and the inability to accurately describe the consistency of participants' transitions through topic space at locations where those transitions frequently occurred in multiple, orthogonal directions (see \textit{Methods} for analysis details).  We formalized this optimization as the combination of hyper-parameter values that satisfied
\[
\argmax_{\left\{K, \gamma, \delta, \tau, \varphi~\in~E~\mathrm{|}~\Phi\left(K, \gamma, \delta, \tau, \varphi \right) \right\}} \left[\mathrm{corr}\left(\Upsilon\left(\xi_{K, \gamma, \delta, \tau, \varphi}\right),~\Upsilon\left(\chi\right)\right)\right],~\mathrm{where}
\]
\[
\Phi\left(K, \gamma, \delta, \tau, \varphi \right) = \argmin_{K, \gamma, \delta, \tau, \varphi} \left[\Gamma\left(\xi_{K, \gamma, \delta, \tau, \varphi}\right)\right],
\]
$\xi$ is the video's trajectory through the manifold space; $\chi$ is the original, 100-dimensional video trajectory; $\Upsilon$ is a function that computes a condensed matrix of pairwise distances between event vectors (computed using correlation distance in the original topic space and Euclidean distance in the manifold space); $\mathrm{corr}\left(\Upsilon\left(\xi\right), \Upsilon\left(\chi\right)\right)$ is the correlation between the sets of pairwise distances, and $\Gamma$ is the number of instances in which lines drawn between consecutive video event embeddings intersected each other.   The sets of hyper-parameter values we searched over comprised: $K \in \{10x \in \mathbb{Z}~|~10 < x <  22\} \cup \{161\}$ (a range roughly centered on half the total number of events, 161, in order to balance representations of local and global structure); $\gamma \in \{1, 3, 5, 7, 9\}$; $\delta \in \{0.1, 0.3, 0.5, 0.7, 0.9\}$; $\tau \in \{x \in \mathbb{Z}~|~0 < x < 101\}$; and $\varphi \in {S\choose3},~\mathrm{where}~S=\{\mathrm{video~events,~average~recall~events, individual~recall~events}\}$.  The optimal parameters (that yielded $\Phi=0$) were $K=170$, $\gamma=7$, $\delta=0.7$, $\tau=41$, with the order of sequence $\varphi$ as the average recall events, video events, and individuals' recall events, vertically concatenated, in order.

@@

\begin{figure}[p!]
\centering
\includegraphics[width=\textwidth]{figs/matchmats}
\caption{\small \textbf{Video-recall event correlation matrices.}  Each panel is in the same format as Figure~2G in the main text.  The yellow boxes mark the matched video event for each recall event (i.e., the maximum correlation in each column).}
\label{fig:matchmats}
\end{figure}

\begin{figure}[p!]
\centering
\includegraphics[width=.9\textwidth]{figs/k_optimization}
\caption{\small \textbf{Video and recall trajectory \textit{K}-optimization functions.}  We selected the optimal $K$-value for the video and each recall trajectory, using the formula described in \textit{Methods}. This computation resulted in a curve for each trajectory, describing the Wasserstein distance between the distributions of within-event and across-event topic vector correlations, as a function of $K$.}
\label{fig:k_optimization}
\end{figure}

>>>>

\begin{figure}[]
\centering
\includegraphics[width=1\textwidth]{figs/feature_value}
\caption{\small \textbf{Feature importance analysis.} \textbf{A.} Contributions of each feature type to the structure of the episode trajectory. The bar heights reflect the correlation between the episode trajectory computed using all features with a episode trajectory computed using all features except the indicated feature.  (Lower bars reflect features that contribute more substantially to the episode trajectory's shape.) \textbf{B.} Which features are preserved during recall?  The bar heights reflect the (average) across-participant correlations between the episode and recall trajectories.  Error bars denote bootstrap-estimated standard error of the mean.  \textbf{C.} Feature correlation matrix.  Each entry displays the correlation between episode topic trajectories created using only the indicated (row/column) features.}
\label{fig:feature-importance}
\end{figure}

We also carried out an analysis of which annotated features tended to shape aspects of the episode topic trajectory's structure that were preserved in participants' recalls.  Specifically, we computed the timepoint-by-timepoint correlation matrix of the episode topic trajectory, and correlated the proximal diagonals from its upper triangle with those of the timepoint-by-timepoint correlation matrices for each participant's recall topic trajectory (resampled using linear interpolation to have the same number of timepoints as the episode trajectory).  This yielded a single correlation coefficient for each participant.  We then repeated this analysis with each annotated feature held out in turn.  Observing a lower correlation between the episode and recall trajectories (when a given feature was held out) would indicate that participants utilize changes in that feature's content to discriminate between sections of the episode when organizing their recalls.  We found that hand-annotated narrative details were the most heavily utilized type of feature, whereas changes in the text present on-screen, the indoor/outdoor distinction between scenes, the camera angle, the names of the various characters on screen, and which character was speaking tended not to impact participants' recall structures (i.e., removing those features resulted in a \textit{greater} episode-recall topic vector correlation than including them; Fig.~\ref{fig:feature-importance}B).

Next, we wondered how the different types of features might relate.  For example, knowing which character is in focus during a given scene may also provide information about which character is speaking.  We computed episode topic trajectories from annotations of each individual feature, in turn, and then compared the proximal diagonals from the upper triangles of the temporal correlation matrices of all pairs of features.  This provided additional confirmation that the shape of the full trajectory (including all types of features) was largely driven by narrative details.  We also found that character-driven features (characters on screen, characters speaking, and characters in focus) were strongly correlated.  Other details, such as the presence or absence of music, led to very different topic trajectories (Fig.~\ref{fig:feature-importance}C).
%\FloatBarrier

\subsection*{Creating a low-dimensional embedding space}
Figures~7 and 8C in the main text display two-dimensional projections of the 100-dimensional topic trajectories for the episode (7A, 8C), average recall (7B), and each individual's recall (7C, 8C).  We created these embeddings using the Uniform Manifold Approximation and Projection algorithm \citep[UMAP;][]{McInEtal18} called from our high-dimensional visualization and text analysis software, \texttt{HyperTools}~\citep{HeusEtal18a}.  An advantage of the UMAP algorithm over comparable manifold learning techniques (such as $t$-SNE) is that UMAP explicitly attempts to preserve the global structure of the data \citep{McInEtal18,BechEtal19} by constructing a space where distance on the manifold is standard Euclidean distance, with respect to the global coordinate system.  This was important in our use case, as we wanted to visualize both the evolving structure of the episode and the spatial relationships between presented and recalled content.

@@

Two other parameters ultimately affect this balance between preserving local and global structure: the seed ($\tau$) for the (pseudo-)random number generator, and the order of the observations (i.e., trajectories) to be embedded ($\varphi$).  As described in \textit{Methods}, we ensured the episode and recall events were projected onto the \textit{same} low-dimensional manifold by fitting the embedding model to a stacked matrix of all episode, average recall, and individuals' recall events.  After initializing the low-dimensional simplicial set (by default, using a spectral embedding of the high-dimensional simplicial set's fuzzy 1-skeleton), UMAP optimizes the embedding using stochastic gradient descent with cross-entropy as a cost function.  During optimization, indices of the data are sampled at random, and thus the local minimum achieved by the optimization is dependent on both the state of the RNG and the sequence in which observations are concatenated.

We performed a grid search over pre-specified values of these hyper-parameters, and optimized the manifold space for visualization based on two criteria: First, that the 2D embedding of the episode trajectory should reflect its original 100-dimensional structure as faithfully as possible. Second, that the path traversed by the embedded episode trajectory should intersect itself a minimal number of times.  The first criteria helps bolster the validity of visual intuitions about relationships between sections of episode content, based on their locations in the embedding space.  The second criteria was motivated by the observed low off-diagonal values in the episode trajectory's temporal correlation matrix (suggesting that the same topic-space coordinates should not be revisited; see Figure~2A in the main text) and the inability to accurately describe the consistency of participants' transitions through topic space at locations where those transitions frequently occurred in multiple, orthogonal directions (see \textit{Methods} for analysis details).  We formalized this optimization as the combination of hyper-parameter values that satisfied
\[
\argmax_{\left\{K, \gamma, \delta, \tau, \varphi~\in~E~\mathrm{|}~\Phi\left(K, \gamma, \delta, \tau, \varphi \right) \right\}} \left[\mathrm{corr}\left(\Upsilon\left(\xi_{K, \gamma, \delta, \tau, \varphi}\right),~\Upsilon\left(\chi\right)\right)\right],~\mathrm{where}
\]
\[
\Phi\left(K, \gamma, \delta, \tau, \varphi \right) = \argmin_{K, \gamma, \delta, \tau, \varphi} \left[\Gamma\left(\xi_{K, \gamma, \delta, \tau, \varphi}\right)\right],
\]
$\xi$ is the episode's trajectory through the manifold space; $\chi$ is the original, 100-dimensional episode trajectory; $\Upsilon$ is a function that computes a condensed matrix of pairwise distances between event vectors (computed using correlation distance in the original topic space and Euclidean distance in the manifold space); $\mathrm{corr}\left(\Upsilon\left(\xi\right), \Upsilon\left(\chi\right)\right)$ is the correlation between the sets of pairwise distances, and $\Gamma$ is the number of instances in which lines drawn between consecutive episode event embeddings intersected each other.   The sets of hyper-parameter values we searched over comprised: $K \in \{10x \in \mathbb{Z}~|~10 < x <  22\} \cup \{161\}$ (a range roughly centered on half the total number of events, 161, in order to balance representations of local and global structure); $\gamma \in \{1, 3, 5, 7, 9\}$; $\delta \in \{0.1, 0.3, 0.5, 0.7, 0.9\}$; $\tau \in \{x \in \mathbb{Z}~|~0 < x < 101\}$; and $\varphi \in {S\choose3},~\mathrm{where}~S=\{\mathrm{episode~events,~average~recall~events, individual~recall~events}\}$.  The optimal parameters (that yielded $\Phi=0$) were $K=170$, $\gamma=7$, $\delta=0.7$, $\tau=41$, with the order of sequence $\varphi$ as the average recall events, episode events, and individuals' recall events, vertically concatenated, in order.

@@

\begin{figure}[p!]
\centering
\includegraphics[width=\textwidth]{figs/matchmats}
\caption{\small \textbf{episode-recall event correlation matrices.}  Each panel is in the same format as Figure~2G in the main text.  The yellow boxes mark the matched episode event for each recall event (i.e., the maximum correlation in each column).}
\label{fig:matchmats}
\end{figure}

\begin{figure}[p!]
\centering
\includegraphics[width=.9\textwidth]{figs/k_optimization}
\caption{\small \textbf{episode and recall trajectory \textit{K}-optimization functions.}  We selected the optimal $K$-value for the episode and each recall trajectory, using the formula described in \textit{Methods}. This computation resulted in a curve for each trajectory, describing the Wasserstein distance between the distributions of within-event and across-event topic vector correlations, as a function of $K$.}
\label{fig:k_optimization}
\end{figure}
